Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-08-19 manuscript/main.tex
fairly complete writeup of methods, including regularization

\DeclareMathOperator{\erf}{erf}

@@

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\begin{equation}
\label{eqn:xivec}
\boldsymbol\xi = C \boldsymbol d,
\end{equation}
where the $(n-1)$ by $(n-1)$ matrix
\[
C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
\]
is constant wrt $\mu$ \emph{and} $\eta$, and
\begin{equation}
\label{eqn:d}
d_j \equiv \int_0^\infty r_j(t)\mu(t)dt
\end{equation}
has linear (nonlinear) dependence on $\mu$ ($\eta$).

@@

We next develop an efficient recursive procedure for computing $C$.
Using \ref{eqn:B}
\begin{align*}
C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
&= W_{b,j} - W'_{b,j}
\end{align*}
Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqns 13-15), and we define $W'$ by
\begin{equation}
\label{eqn:W'}
W'_{b,j} = \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
\end{equation}
\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
\[
A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
\]
so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W'} is a definite sum over geometric terms.
Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following second-order recursion in $j$:
\begin{align*}
&W'_{b,2} = 0\\
&W'_{b,3} = \frac{20 (n-2)}{(n+1)(n+2)}\\
&W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
\end{align*}
This completes the framing of mutation rate inference as a linear inverse problem.

@@

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $\ell$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{\ell-1}, t_\ell)$ where $0=t_0 < t_1 < \dots < t_{\ell-1} < t_\ell=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
Let $(y_1,\dots,y_\ell)$ denote the constant population size $eta(t)$ during each epoch, and let $(z_1,\dots,z_\ell)$ denote the constant mutation rate $mu(t)$ during each epoch.
Let $x_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_\ell$, and $x_0\equiv 1$.
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & x_1                & \hdots & \prod_{i=1}^{\ell-1}x_i               \\
1       & x_1^3              & \hdots & \prod_{i=1}^{\ell-1}x_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & x_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{\ell-1}x_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
y_1  &      &        &             &       \\
-y_1 & y_2  &        &             &       \\
     & -y_2 & y_3    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -y_{\ell-1} & y_\ell
\end{bmatrix}.
\end{equation}
Note that the $(n-1)$ by $\ell$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $x_l$ are nonlinear functions of $\boldsymbol y$.

@@

% DEPRECATED:
%
% \subsection*{Constant MLE}\label{sec:model:constant}
%
% We initialize by fitting a constant population size.
% According to WSD's scribbles, the MLE assuming $\eta(t) = \eta_0$ (constant) is $\hat \eta_0 = \frac{S}{2 r H_{n-1}}$, where $S$ is the number of segregating sites (the sum of the observed SFS vector), $r$ is the mutation rate per genome per generation, and $H_{n-1}$ is the $n$th harmonic number.
% This was derived by using the well-known result (cited in Rosen et al.) that the expected SFS for a constant population is given by $\xi_i = \frac{2\eta_0}{i}$.
% Then the likelihood for $\eta_0$ is a Poisson random field parameterized by the $\xi_i$.
%
%
% \subsection*{The coalescent horizon: tempora incognita (see what I did there? ;))}\label{sec:model:horizon}
%
% The coalescent tree height determines the time scale within which the SFS contains demographic information.
% The expected height for n samples and a constant history $\eta(t)=\eta_0$ is the sum of the independent exponentially-distributed intercoalescent expectations:
% \[
% \mathbb{E}[T_n] = \sum_{k=2}^n \frac{\eta_0}{\binom{k}{2}} = 2\eta_0\left(1-\frac{1}{n}\right),
% \]
% where the last equality follows by recognizing the telescoping series.
% Likewise for the variance:
% \[
% \text{Var}[T_n] = \sum_{k=2}^n \left(\frac{\eta_0}{\binom{k}{2}}\right)^2 = \frac{4 \eta _0^2 \left(-6 n^2 \psi ^{(1)}(n)+\left(\left(\pi ^2-9\right) n+6\right) n+3\right)}{3 n^2},
% \]
% where $\psi ^{(1)}(n)$ is the derivative of the digamma function \url{https://en.wikipedia.org/wiki/Trigamma_function}.
%
%
% \subsection*{Asymptotically constant boundary condition}\label{sec:model:boundary}
%
% % Kelley's eqn
% In real data (or realistic simulations) the time region where the SFS has lost information is determined by the expected number of coalescent trees $\rho\ell$, where $\rho$ is the recombination rate and $\ell$ is the genome size.
% For $t$ such that the expected number of trees that have $T_{\text{MRCA}} > t$ becomes less that one, the SFS does not contain information beyond $t$.
% To wit,
% \[
% \mathbb{P}\left(T_{\text{MRCA}}>t\right) < \frac{1}{\rho \ell}.
% \]
% The P\&K model doesn't model finite genome size, so it makes sense to use the CDF of $T_{\text{MRCA}}$ in that setting.

>>>>

\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\lngamma}{\ln\!\Gamma}

@@

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\begin{equation}
\label{eqn:xivec}
\boldsymbol\xi = C \boldsymbol d,
\end{equation}
where the $(n-1)\times(n-1)$ matrix
\[
C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
\]
is constant wrt $\mu$ \emph{and} $\eta$, and
\begin{equation}
\label{eqn:d}
d_j \equiv \int_0^\infty r_j(t)\mu(t)dt
\end{equation}
has linear (nonlinear) dependence on $\mu$ ($\eta$).

@@

We next develop an efficient recursive procedure for computing $C$.
Using \ref{eqn:B}
\begin{align*}
C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
&= W_{b,j}^{(1)} - W_{b,j}^{(2)},
\end{align*}
where
\begin{align}
\label{eqn:W1}
W_{b,j}^{(1)} &\equiv \sum_{k=2}^j k p_{n,k}(b) A_{k,j}\\
\label{eqn:W2}
W_{b,j}^{(2)} &\equiv \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
\end{align}
\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
\[
A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
\]
so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W1} and \ref{eqn:W2} are definite sums over geometric terms.
Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following procedurally generated second-order recursions in $j$:
\begin{align*}
&W_{b,2}^{(1)} = \frac{6}{(n+1)}\\
&W_{b,3}^{(1)} = \frac{10(5n-6b-4)}{(n+2)(n+1)}\\
&W_{b,j+2}^{(1)} = -\frac{(2 j+3) \left(-(2 j-1) W_{b,j+1}^{(1)}  \left(2 j (j+1) \left(b^2 \left(j^2+j-2\right)-6 b-j (j+1)-2\right)-j (j+1) n \left(3 b \left(j^2+j+2\right)+j^2+j-2\right)+\left(j (j+1) \left(j^2+j+6\right)+4\right) n^2+4 n\right)-(j-1) (j+1)^2 (j-n) W_{b,j}^{(1)}  (4 (n+1)-j (j+2) (b-n-1))\right)}{j^2 (j+2) (2 j-1) (j+n+1) \left(-b j^2+b+\left(j^2+3\right) (n+1)\right)}
\end{align*}
and
\begin{align*}
&W_{b,2}^{(2)} = 0\\
&W_{b,3}^{(2)} = \frac{20 (n-2)}{(n+1)(n+2)}\\
&W_{b,j+2}^{(2)} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W_{b,j}^{(2)}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W_{b,j+1}^{(2)}\right)
\end{align*}
This completes the framing of mutation rate inference as a linear inverse problem.

@@

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_m$, and $u_0\equiv 1$.
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
y_1  &      &        &             &       \\
-y_1 & y_2  &        &             &       \\
     & -y_2 & y_3    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -y_{m-1} & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

@@

\subsection{Tempora incognita: regularization and the coalescent horizon}\label{sec:model:loss}

In linkage equilibrium the log likelihood function of the history $(\boldsymbol y, \boldsymbol z)$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ is the Poisson random field \citep{}
\begin{align}
\ell(\boldsymbol y, \boldsymbol z) &= \sum_{b=1}^{n-1}\left(x_b\log(\xi_b) - \xi_b - \log(x_b!)\right)\\
&= \boldsymbol x^\intercal\log(\boldsymbol\xi) - \boldsymbol 1^\intercal\left(\boldsymbol\xi + \lngamma(\boldsymbol x + \boldsymbol 1)\right)\\
&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal\left(C M(\boldsymbol y) \boldsymbol z + \lngamma(\boldsymbol x + \boldsymbol 1)\right),
\end{align}
where $\log(\cdot)$ and the log-gamma function $\lngamma(\cdot)$ operate on vectors elementwise.

The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
To penalize oscillatory solutions we introduce regularizers on the history derivatives.
Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
At times for which the CDF is near unity, the TMRCA has probably occurred, so we should penalize heavily because the data are not informative about the history at this time.
At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
From \eqref{eqn:r} the CDF $1 - r_2(t)$, so we define the regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ as
\begin{align}
\rho_\eta[\eta(t)] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)\left(1-r_2(t)\right)dt,\\
\rho_\mu[\eta(t), \mu(t)] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)\left(1-r_2(t)\right)dt.
\end{align}
With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and it is straightforward to show (using notation as in section \ref{sec:model:pcsws}) that these become
\begin{align}
\rho_\eta(\boldsymbol y) &= \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2,\\
\rho_\mu(\boldsymbol y, \boldsymbol z) &= \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2.
\end{align}
where
\begin{equation}
\label{eqn:Gamma}
\Gamma(\boldsymbol y) \equiv
\begin{bmatrix}
\sqrt{1-u_0} &     &        &                       \\
  & \sqrt{1-u_0 u_1} &        &                       \\
  &     & \ddots &                       \\
  &     &        & \sqrt{1-\prod_{i=1}^{m-1}u_i}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}.
\end{equation}
Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.

Our regularized loss function is then
\begin{align}
L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z)\\
&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2
\end{align}
where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.
